---
title: "NBA Clustering Report"
author: "Quinn Glovier"
date: "`r Sys.Date()`"
output: rmdformats::downcute
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, include=FALSE}
#Required Packages
library(tidyverse)
library(plotly)
library(htmltools)
library(devtools)
library(caret)
library(NbClust)
```

### Report

The goal of my data analysis was to find players that had high performance statistics compared to their salary on average. To classify by salary, I grouped the players into "Above Average", "Average", and "Below Average" based on the bottom 25% and top 75% thresholds for salaries. I chose to measure performance were Games Played, Games Started, Minutes Played, Field Goals, 3 Point Shots, 2 Point Shots, Free Throws, Total Rebounds, Assists, Turnovers, and Points Total.

When running the model, I grouped the data into three clusters, with Cluster 1 having the highest performing players and Cluster 3 having the lowest performing players. When evaluating the data with the "Elbow Chart" method and the "NBClusters" method, I found that 2 and 3 clusters were the most reccomended numbers, but 3 performed better. 

When evaluating the results, the 3D graph was most helpful to me. I chose to review the players based on Minutes Played (shows how valuable of a player they are), 3 Point Shots (an offensive statistic), and Turnovers (a defensive statistic), so that the review process was more holistic and looked at different performance aspects. I selected below-average and average paid players in Cluster 1 to find high performing players with lower salaries. The players I picked are:


1. Jakob Poeltl: This player had  around 29 minutes of play time, which was shorter than some of the other high-performing players in the average salary group. He had 6 2-point shots and 1.6 turnovers, showing strong offensive capabilities.


2. Cade Cunningham: This player was also within the average salary group and played on average around 32.6 minutes. He had 4.9 for 2 Point Shots and 3.7 Turnovers, which was one of the highest turnover scores, showing a strong defensive ability.


3. Herbert Jones: This was one of, if not the best player in the below-average salary category, meaning he would probably be less expensive to recruit compared to the other two players. He had an average playing time of around 29.9 minutes, 2.3 2 Point Shots, and 1.8 Turnovers. Looking at his stats in the original dataset, Jones also had 3.8 Total Rebounds, and 9 Total Points, showing potential.


```{r, include = FALSE}
#Data Cleaning

#Loading the data
salaries <- read_csv("nba_salaries_22.csv")
perf <- read_csv("NBA_Perf_22.csv")
View(salaries)
View(perf)

#Merging the data
nba <- inner_join(salaries, perf)
view(nba)

#Removing the NA values
summary(is.na(nba))
nba <- nba[complete.cases(nba),]
summary(is.na(nba))

#Remove Duplicates
nba$Salary <- gsub("[^[:alnum:]]", "", nba$Salary)
nba$Salary <- as.numeric(nba$Salary)
nba <- group_by(nba, Player)%>%
  summarise_each(funs(mean))
View(nba)
nba <- nba[, -c(3,5)]
str(nba)


#Creating classifcations of players by salary
#First, dividing the table by 100, since removing the special characters added the "cents" of the salary as dollars
nba$Salary = nba$Salary/100
#Splitting the data into three categories based on the salary. Players in the middle 50% are classified as having an "average" salary, while the bottom 25% is below average, and the top 25% is above average.
fivenum(nba$Salary)
salary_labels <- c("Below Average", "Average", "Above Average")
nba <- mutate(nba, S_Rating = case_when(
  nba$Salary <= 2101800 ~ salary_labels[1],
  nba$Salary >2101800 & nba$Salary<= 13545000 ~ salary_labels[2],
  nba$Salary > 13545000 ~ salary_labels [3]
))
view(nba)


#Select the variables to be included in the cluster 
#Choosing Games Played, Total Rebounds, and 3-Point Shots
nba1 <- nba[, c("G", "GS", "MP", "FG", "3P", "2P", "FT", "TRB", "AST", "TOV", "PTS")]
view(nba1)
normalize <- function(x){
 (x - min(x)) / (max(x) - min(x))
}

nba1 <- normalize(nba1)
view(nba1)
```

```{r, include = FALSE}
set.seed(2020)
kmeans_obj_nba = kmeans(nba1, centers = 3, 
                        algorithm = "Lloyd")

kmeans_obj_nba
```

### 2D Graph Showing how Minutes Played and 3 Point Shots Correlate to Salary

```{r, echo = FALSE, eval = TRUE}
#2D Graph
clusters_nba= as.factor(kmeans_obj_nba$cluster)

nba$S_Rating <- as.factor(nba$S_Rating)
ggplot(nba, aes(x = MP, 
                            y = `3P`,
                            color = S_Rating,  
                            shape = clusters_nba)) + 
  geom_point(size = 6) +
  ggtitle("Minutes Played vs 3 Point Shots in the NBA classified by Salary") +
  xlab("Minutes Played") +
  ylab("Count of 3 Point Shots") +
  scale_shape_manual(name = "Cluster", 
                     labels = c("Cluster 1", "Cluster 2", "Cluster 3"),
                     values = c("1", "2", "3")) +
  scale_color_manual(name = "Salary",
                     labels = c("Above Average", "Average", "Below Average"),
                     values = c("#2FBF71", "#8EE3F5", "#DD403A")) +
  theme_light()
```

### 3D Graph Showing How Minutes Played, 3 Point Shots, and Turnovers Affect Salary

```{r, echo = FALSE, eval = TRUE}
nba2 <- mutate(nba, clusters = clusters_nba)
view(nba2)

fig <- plot_ly(nba2, 
               type = "scatter3d",
               mode="markers",
               symbol = ~clusters,
               x = ~MP, 
               y = ~`3P`, 
               z = ~`TRB`,
               color = ~S_Rating,
               colors = c("#2FBF71", "#8EE3F5", "#DD403A"),
               alpha = 0.8,
               text = ~paste('Player:', Player))%>%
  layout(title = "Minutes Played, 3 Point Shots, and Total Rebounds")


fig
```
### "Elbow" Graph Reccomending the Number of Clusters to be Used

```{r, echo = FALSE, eval= TRUE}

explained_variance <- function(data_in, k){
  
  # Running the kmeans algorithm.
  set.seed(1)
  kmeans_obj = kmeans(data_in, centers = k, algorithm = "Lloyd", iter.max = 30)
  
  # Variance accounted for by clusters:
  # var_exp = intercluster variance / total variance
  var_exp = kmeans_obj$betweenss / kmeans_obj$totss
  var_exp  
}



explained_var_nba <-sapply(1:10, explained_variance, data_in = nba1)

elbow_data_nba <- data.frame(k = 1:10, explained_var_nba)

ggplot(elbow_data_nba, 
       aes(x = k,  
           y = explained_var_nba)) + 
  geom_point(size = 4) +
  geom_line(size = 1) + 
  xlab('k') + 
  ylab('Inter-cluster Variance / Total Variance') + 
  theme_light()

```



```{r, include = FALSE}
nbclust_obj_nba = NbClust(data = nba1, method = "kmeans")

nbclust_obj_nba

```

### Graph Showing What Number Of Clusters is best for analyzing the data
```{r, echo = FALSE, eval = TRUE}
freq_k_nba = nbclust_obj_nba$Best.nc[1,]
freq_k_nba = data.frame(freq_k_nba)

ggplot(freq_k_nba,
       aes(x = freq_k_nba)) +
  geom_bar() +
  scale_x_continuous(breaks = seq(0, 15, by = 1)) +
  scale_y_continuous(breaks = seq(0, 12, by = 1)) +
  labs(x = "Number of Clusters",
       y = "Count",
       title = "Cluster Analysis")
```

